---
title: "APE Package"
author: "Avery Shepherd"
date: "5/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


The <code>ape</code> package (Analysis of Phylogenetics and Evolution) is used within the community of evolutionary biologists for data analysis and as a framework for the development of new analytical methods.
```{r}
library(ape)
```

To read and load trees into R using ape, you can use the <code>read.nexus</code> or <code>read.tree</code> functions. The <code>read.nexus</code> function reads NEXUS formatted files while the <code>read.tree</code> function reads Newick formatted trees.
```{r}
tree <- read.nexus("./Potential Nexus Files/T6506.nex")
tree
```

When the tree is read in and processed, it creates a list of class phylo

```{r}
class(tree)
str(tree)
```

The <code>edge</code> variable in the tree contains the beginning and ending node number for the nodes and tips in the tree
```{r}
head(tree$edge)
```

The <code>tip.label</code> variable in the tree contains  the labels for the tips in the tree
```{r}
head(tree$tip.label)
```

<code>Nnode</code> contains the number of internal nodes in the tree
```{r}
head(tree$Nnode)
```

Graphical output in form of a cladogram can be obtained by calling <code>plot</code> with the tree as a parameter. This calls the <code>plot.phylo</code> function of the <code>ape</code> package. 
```{r}
plot(tree, cex = 0.5)
```



```{r}
groupInfo <- split(tree$tip.label, 
                   gsub("_\\w+", "", tree$tip.label))
head(groupInfo)
```

# this should work.... but my code is not working with ggtree and i cannot diagnose so rip
```{r}
a <- groupOTU(tree, groupInfo)
ggtree(a, aes(color = group), layout='circular')
```

```{r}

```


```{r}

```




