---
title: "Phytools"
author: "Emily Lessig"
date: "5/10/2021"
output: html_document
---

'phytools': Phylogenetic Tools for Comparative Biology (and Other Things)


phytools takes advantage of functionality developed in other packages such as ape for many types of phylogenetic tree input and manipulation

Examples of phytools to demonstrate its functions:

```{r}

library(readxl)
library(ape)
library(phytools)

```

'phylo.heatmap': creates a multivariate phylogenetic heatmap 

https://rdrr.io/cran/phytools/man/phylo.heatmap.html

```{r}
#Run through example using datasets in R

## load data
data(anoletree)
data(anole.data)
## compute phylogenetic residuals
anole.data<-as.matrix(anole.data)
anole.resids<-cbind(anole.data[,1],
    phyl.resid(anoletree,anole.data[,1,drop=FALSE],
    anole.data[,2:ncol(anole.data)])$resid)
colnames(anole.resids)[1]<-"SVL"
## plot phylogenetic heatmap
phylo.heatmap(anoletree,anole.resids,
    split=c(0.7,0.3),fsize=c(0.4,0.8,0.8),
    standardize=TRUE)
par(mar=c(5.1,4.1,4.1,2.1)) ## reset margins to default
```

'phyl.pca': performs phylogenetic PCA 

https://rdrr.io/cran/phytools/man/phyl.pca.html

```{r}
data(anoletree)
data(anole.data)
## run phylogenetic PCA
anole.pca<-phyl.pca(anoletree,anole.data)
print(anole.pca)
## plot results
plot(anole.pca)
biplot(anole.pca)

```
'phylo.to.map': creates an object of class "phylo.to.map" and plots that object (where tips of the tree point to coordinates on a geographic map). Recent update also allows to mape more than one geographic location per taxon.

```{r}
#Example using simulated data

install.packages("mapdata")
library(mapdata)

World.tree<-pbtree(n=26,scale=100)
World.tree$tip.label<-replicate(Ntip(World.tree),
    paste(sample(LETTERS,1),".",
    paste(sample(letters,round(runif(n=1,min=3,max=10))),
    collapse=""),
    sep=""))
lat<-fastBM(World.tree,sig2=10,bounds=c(-90,90))
long<-fastBM(World.tree,sig2=80,bounds=c(-180,180))
World<-cbind(lat,long)
for(i in 1:Ntip(World.tree)){
    ni<-sample(0:2,1)
    for(j in 1:ni){ 
        World<-rbind(World,c(World[i,1]+rnorm(n=1,sd=4),
            World[i,2]+rnorm(n=1,sd=4)))
        rownames(World)[nrow(World)]<-rownames(World)[i]
    }
}

head(World, 20)

obj <- phylo.to.map(World.tree, World, plot=FALSE)

plot(obj, ftype="i")

cols<-setNames(sample(rainbow(n=Ntip(World.tree))),
    World.tree$tip.label)
plot(obj,colors=cols,ftype="i",fsize=0.6,cex.points=c(0.7,1.2))
```

'phylANOVA': performs simulated based phylogenetic ANOVA and can conduct post-hoc comparisons of means among groups

https://rdrr.io/cran/phytools/man/phylANOVA.html

```{r}

#simulate the data:

tree<-pbtree(n=100)
Q<-matrix(c(-2,1,1,
    1,-2,1,
    1,1,-2),3,3)
rownames(Q)<-colnames(Q)<-letters[1:3]
x<-as.factor(sim.history(tree,Q)$states)
y1<-fastBM(tree)
y2<-fastBM(tree,sig2=0.5)+as.numeric(x)

phenogram(tree,y1,ftype="off",col=make.transparent("blue",0.5))
tiplabels(pie=to.matrix(x,letters[1:3]),
    piecol=colorRampPalette(c("blue", "yellow"))(3),
    cex=0.4)
    
phenogram(tree,y2,ftype="off",col=make.transparent("blue",0.5))
tiplabels(pie=to.matrix(x,letters[1:3]),
    piecol=colorRampPalette(c("blue", "yellow"))(3),
    cex=0.4)
    
phylANOVA(tree,x,y1)

phylANOVA(tree,x,y2)

```


