---
title: "Phylogenetic Comparative Methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Put the title of your vignette here}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
  
# Phylogenetic Comparative Methods

## Introduction to phylogenies
Phylogenetic trees are branching diagrams illustrating the evolutionary relationships among taxa. 

* Each **node** represents a speciation event in evolution.

* A **branch** connects two nodes of the tree

* **Terminal Nodes** represent the data under comparison. Also known as **Operational Taxonomic Units** *(OTUs)*

# Introduction to Phylogenetic comparative methods (PCMs)
PCMs enable us to study the history of organismal evolution and diversification. With it we can explore in depth these topics:

* Evolutionary ties between organisms

* Find and understand relationships between an ancestral sequence and its descendants

* Estimate time of divergence between a group of organisms that share a common ancestor

## <code>APE</code> package and the <code>phylo</code> class
The <code>ape</code> package (Analysis of Phylogenetics and Evolution) is used within the community of evolutionary biologists for data analysis and as a framework for the development of new analytical methods.
```{r}
library(ape)
```

To read and load trees into R using ape, you can use the <code>read.nexus</code> or <code>read.tree</code> functions. The <code>read.nexus</code> function reads NEXUS formatted files while the <code>read.tree</code> function reads Newick formatted trees.
```{r}
tree <- read.nexus("./Potential Nexus Files/T6506.nex")
tree
```

When the tree is read in and processed, it creates a list of class phylo. This phylo type is the basis of most PCMs. 
```{r}
class(tree)
str(tree)
```

The <code>edge</code> variable in the tree contains the beginning and ending node number for the nodes and tips in the tree
```{r}
head(tree$edge)
```

The <code>tip.label</code> variable in the tree contains  the labels for the tips in the tree
```{r}
head(tree$tip.label)
```

<code>Nnode</code> contains the number of internal nodes in the tree
```{r}
head(tree$Nnode)
```

Graphical output in form of a cladogram can be obtained by calling <code>plot</code> with the tree as a parameter. This calls the <code>plot.phylo</code> function of the <code>ape</code> package. 
```{r}
plot(tree, cex = 0.5)
```

## Phytools package
The <code>phytools</code> package (Phylogenetic Tools for Comparative Biology (and Other Things)) takes advantage of functionality developed in other packages such as <code>ape</code> for many types of phylogenetic tree input and manipulation 

```{r}
library(phytools)
```

The <code>phylo.heatmap</code> function creates a multivariate phylogenetic heatmap

```{r}
library(maps)
```

To read and load trees into R for <code>phytools</code> for this function, you can use <code>anoletree</code> which is a phylogeny of Greater Antillean anole species with a mapped discrete character - ecomorph class and <code>anole.data</code> which is a data frame of morphological characters (both data and tree are from Mahler et al. 2010)

```{r}
data("anoletree")
data("anole.data")
```

Using <code>as.matrix</code> you can convert the anole data table into a matrix 

```{r}
anole.data <-as.matrix(anole.data)
```

Next you can compute the phylogenetic residuals. 

```{r}
anole.data<-as.matrix(anole.data)
anole.resids<-cbind(anole.data[,1],
    phyl.resid(anoletree,anole.data[,1,drop=FALSE],
    anole.data[,2:ncol(anole.data)])$resid)
colnames(anole.resids)[1]<-"SVL"

```

You can plot the phylogenetic heatmap using <code>phylo.heatmap</code>

```{r}
phylo.heatmap(anoletree,anole.resids,
    split=c(0.7,0.3),fsize=c(0.4,0.8,0.8),
    standardize=TRUE)
```

The <code>phyl.pca</code> function performs phylogenetic PCA

You can use the same anole data set from Mahler et al. 

```{r}
data(anoletree)
data(anole.data)
```

Using <code>phyl.pca</code> allows you to run the phylogenetic PCA 

```{r}
anole.pca <- phyl.pca(anoletree, anole.data)
```

<code>print</code> and <code>summary</code> allows you to see the results of the PCA (this shows you the loadings of the various PC's and proportion of variance)

```{r}
print(anole.pca)
summary(anole.pca)
```

<code>plot</code> gives you a screeplot. This shows you how much variation is explained by each component (each PC).
```{r}
plot(anole.pca)
```

To visualize the loadings, you can use <code>biplot</code>

```{r}
biplot(anole.pca)
```

'phylo.to.map' creates an object of class "phylo.to.map" and plots that object (where tips of the tree point to coordinates on a geographic map). It is possible to map more than one geographic location per taxon.

```{r}
library(mapdata)
```

We can simulate a phylogenetic tree to plot with <code>pb.tree()</code>.
   ?what are the units of scale?
   ?do we need a citation for this simulation code or did Emily write it all?
```{r}
# 10 species, total length of the tree is 100
World.tree<-pbtree(n=10,scale=100)
# setting labels for branches
World.tree$tip.label<-replicate(Ntip(World.tree),
    paste(sample(LETTERS,1),".",
    paste(sample(letters,round(runif(n=1,min=3,max=10))),
    collapse=""),
    sep=""))
# simulating latitudes and longitudes
lat<-fastBM(World.tree,sig2=10,bounds=c(-90,90))
long<-fastBM(World.tree,sig2=80,bounds=c(-180,180))
# group data
World<-cbind(lat,long)
for(i in 1:Ntip(World.tree)){
    ni<-sample(0:2,1)
    for(j in 1:ni){ 
        World<-rbind(World,c(World[i,1]+rnorm(n=1,sd=4),
            World[i,2]+rnorm(n=1,sd=4)))
        rownames(World)[nrow(World)]<-rownames(World)[i]
    }
}

```

We can see the data we've simulated here.
```{r}
head(World, 20)
```

Use <code>phylo.to.map()</code> to display the geographic locations associated with each species.
```{r}
obj <- phylo.to.map(World.tree, World, plot=FALSE)
# default settings
plot(obj, ftype="i")

# colors, font size, point size, etc. can be customized
cols<-setNames(sample(rainbow(n=Ntip(World.tree))),
    World.tree$tip.label)
plot(obj,colors=cols,ftype="i",fsize=0.8,cex.points=c(0.7,1.2))
```

Next, we will look at <code>phylANOVA()</code> which performs simulation based phylogenetic ANOVA and can conduct post-hoc comparisons of means among groups.

```{r}
# simulate data for the analysis
tree<-pbtree(n=100)
Q<-matrix(c(-2,1,1,
    1,-2,1,
    1,1,-2),3,3)
rownames(Q)<-colnames(Q)<-letters[1:3]
x<-as.factor(sim.history(tree,Q)$states)
y1<-fastBM(tree)
y2<-fastBM(tree,sig2=0.5)+as.numeric(x)
```

```{r}
phenogram(tree,y1,ftype="off",col=make.transparent("blue",0.5))
tiplabels(pie=to.matrix(x,letters[1:3]),
    piecol=colorRampPalette(c("blue", "yellow"))(3),
    cex=0.4)
    
phenogram(tree,y2,ftype="off",col=make.transparent("blue",0.5))
tiplabels(pie=to.matrix(x,letters[1:3]),
    piecol=colorRampPalette(c("blue", "yellow"))(3),
    cex=0.4)
```

```{r}
phylANOVA(tree,x,y1)

phylANOVA(tree,x,y2)
```

